Bayesian Reconstruction of Coevolutionary Histories

### 2013-09-29 executive summary/article.tex
Final draft sent to siemens

\section*{Abstract}

Symbiotic interactions are a fundamental aspect of life and understanding how these relationships coevolve remains an important goal in biology. Phylogenetic trees show the ancestral relationships between a group of organisms and can be used to learn about evolutionary processes. After constructing a phylogeny for a host and symbiont organism from molecular sequences, we can create a map of ancestral symbionts to ancestral hosts called a reconciliation. The reconciliation often reveals discrepancies between the two trees which we can explain with the biological events duplication, host-switch, and loss. Current approaches to this problem apply a penalizing score to each event and optimize the reconciliation between the two trees to reduce these events. I instead assign each event a rate and develop an algorithm to calculate the probability of a symbiont tree given a host tree, a reconciliation, and the event rates and implement it in a Bayesian MCMC framework where all parameters are simultaneously estimated or integrated out. I test my algorithm on data simulated under the same coevolutionary model and find good, but not perfect, reconstruction of coevolutionary histories, as well as a flaw in the operator. Finally, I discuss potential extensions for my model, particularly to coevolutionary theory.

>>>>

\section*{Executive Summary}

Whether an animal, plant, bacterium, or even a virus, most organisms depend on interactions with other species. Understanding how these interactions emerge and change is an important goal in biology. This process is called coevolution because multiple interacting organisms are evolving together, generally a host organism and a symbiont organism. Similar to constructing a family tree, and observing how traits such as eye- and hair-color are passed from one generation to the next, we can construct family trees of organisms from their DNA to learn how their traits evolve over time. Unlike our knowledge of our parents and grandparents, whose eye- and hair-colors we know, we do not know what the ancestral organisms were like. One solution to this problem is to make a guess about the ancestors' traits, and try to use it to explain the traits that we see today. For organismic interactions, we can guess which ancestral host was partnered with which ancestral symbiont and try to explain their current partnerships with biological events. Examples of these events are host-switches, when an organism changes its partner. I create an algorithm to calculate the probability of a guess using these events, so that by making lots of guesses with a computer we can reconstruct the ancestral partnerships between a host and symbiont organism. I test my algorithm on simulated data and show that it can find a good, but not perfect, reconstruction of these ancestral partnerships.
